```{r}
#| label: setup-p1
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", dev = "svglite", dev.args = list(bg = "transparent"))

library(DoseFinding)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(20260813)

theme_set(
  theme_minimal(base_size = 15) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "right",
          plot.background = element_rect(fill = "transparent", colour = NA),
          panel.background = element_rect(fill = "transparent", colour = NA))
)
```

# Welcome :wave: {background-color="#1d5c4a"}

## Get set up :hammer_and_wrench: {.small}

::: {.fragment .keybox}

**Start this now.** Downloading and installing takes a few minutes, so kick it
off while we do introductions.

:::


::: {.fragment}

Parts 2 and 3 are hands-on. You will get far more out of them with R open and
the project already downloaded. The full walkthrough — including a no-install
hosted fallback — is in [Setup](#setup).

:::


## Your instructors :busts_in_silhouette: {.medium}

::: {.columns}
::: {.column width="50%"}
::: {.bio-card}

**Mitch Thomann, PhD** — Parts 1 & 3



- PhD in Biostatistics, University of Iowa (2015); thesis on novel dose-finding designs
- 10+ years designing trials with R-based simulation across two sponsors
- Currently Head of Modeling and Simulation at Viridian Therapeutics
- Focus areas: Bayesian trial design, dose-finding, Probability of Success


:::
:::

::: {.column width="50%"}
::: {.bio-card}

**Saumil Shah, PhD** — Part 2



- PhD in Mathematical Biology, Max Planck Institute, Germany (2024)
- Thesis: expediting translational research using quantitative methods
- Currently in Simulation Hub at Boehringer Ingelheim Pharma
- Focus: high-throughput operational and statistical simulation of clinical trials


:::
:::

:::

## Housekeeping :house:

- **Three hours, three parts**, **1:30 PM – 5:00 PM (placeholder, per RISW 2026)**,
  with **two 15-minute breaks**
- **Parts 1 and 2 are the backbone** — the framework, and how to write it in R
- **Part 3 is deliberately flexible.** It is hands-on and interactive, and how
  far we take it depends on the room
- Part 2 and Part 3 both go better with R open
- Questions welcome throughout — this is a small enough room for it


## Agenda :spiral_calendar: {.small}

| | Time | Session | | |
|:--:|---|---|---|---|
| **1** | 1:30 – 2:30 PM | **Overview of Trial Simulations**<br/>*Mitch · 60 min* | Motivation · modern capabilities · regulatory guidance<br/>The five essential components, on one running example | **← you are here** |
| | 2:30 – 2:45 PM | 15 minute break | | |
| **2** | 2:45 – 3:45 PM | **Programming Key Elements in R**<br/>*Saumil · 60 min* | Objectives · scenarios · data generation<br/>Complex designs · evaluation · parallelization | *after the break* |
| | 3:45 – 4:00 PM | 15 minute break | | |
| **3** | 4:00 – 5:00 PM | **Hands-on Exercise**<br/>*Mitch · flexible* | Build one design end to end<br/>Compare designs across the full study | *after the break* |


## What we expect from today :dart: {.small}

::: {.columns}
::: {.column width="52%"}

**We do expect**

- A shared, concrete understanding of what a trial simulation **is** and what
  it is **for**
- Fluency with the five components every simulation has — and the vocabulary
  to discuss them
- Enough R-level detail to go home and build one
- Engagement: this framework improves when it meets other people's problems

:::
::: {.column width="48%"}

**We don't expect**

- That you leave an expert
- That you adopt any one package wholesale

:::
:::

## A note on what follows :speech_balloon:

::: {.fragment}

Everything in this course — the framework, the terminology, the specific
recommendations — is built from years of hands-on trial simulation work across
multiple organizations.

:::


::: {.fragment .keybox}

These are **our professional opinions**, not official recommendations or positions of
our employers or any regulatory authority. Where regulatory guidance is directly
relevant, we cite and quote it; everything beyond that is judgment we've formed through
practice — and are glad to defend, but which you should weigh as one experienced
perspective, not policy.

:::


# Motivation :fire: {background-color="#1d5c4a"}

## Real life is a single shot on goal :soccer:

- In a clinical trial we effectively get **one** shot at a chosen design.
- The design is locked before a single patient enrolls, and the consequences of
  getting it wrong are measured in years and patients.
- Simulation is the practice gym: we can play the trial thousands of times under
  plausible realities before we commit.


::: {.fragment}

> "In my opinion, it is more desirable to kill computer-generated patients, rather
> than real ones, in order to calibrate design parameters."
>
> — Peter Thall [@thall2015smart]

:::


## The familiar workflow :arrows_counterclockwise:

::: {.fragment}

```{mermaid}
%%| fig-width: 11
%%{init: {"themeVariables": {"fontSize": "18px", "edgeLabelBackground": "#f6faf8"}, "flowchart": {"htmlLabels": true, "useMaxWidth": false}}}%%
flowchart LR
  A(Trial design options) -- Closed-form tool --> B(Operating characteristics) --> C(Decide)
```

:::


::: {.fragment}

```{mermaid}
%%| fig-width: 11
%%{init: {"themeVariables": {"fontSize": "18px", "edgeLabelBackground": "#f6faf8"}, "flowchart": {"htmlLabels": true, "useMaxWidth": false}}}%%
flowchart LR
  D(New design option) -- Tool does not fit --> E(Approximate) --> F(Find another tool) --> G(Approximate a bit more)
```

:::

::: {.fragment .zoom}

Every design question that falls outside the tool becomes a negotiation about what
we are *willing to approximate*.

:::


## There is a better way :bulb:

::: {.fragment}

Start in R.

:::


::: {.fragment}

```{mermaid}
%%| fig-width: 11
%%{init: {"themeVariables": {"fontSize": "18px", "edgeLabelBackground": "#f6faf8"}, "flowchart": {"htmlLabels": true, "useMaxWidth": false}}}%%
flowchart LR
  A(Design + truth scenarios) -- Write R code --> B(Operating characteristics) -- Evaluate --> C(Interpret)
  C -- Ask the next question --> A
```

:::


::: {.fragment .keybox}

The goal is not "an R script that answers this question." It is R code that is
**simple, adaptable and flexible** enough to answer the *next* question — because
there is always a next question.

:::


## Why now? :hourglass_flowing_sand: {.medium}

::: {.incremental}
- **Design complexity has outrun closed-form methods**
  - adaptive elements, Bayesian borrowing, multiple and composite endpoints,
    estimand-aligned intercurrent event handling
- **Patient-level generation has become the norm**
  - replaces summary-statistic simulation — closer to how trials actually unfold
- **Compute is no longer the binding constraint**
  - the constraint now is how quickly you can express a new design in code
- **Regulators increasingly expect it** — more on this shortly

:::


## Trial simulation *is* :white_check_mark:

- Repeated virtual trials
  - typically 1,000s–100,000s of them
- Built from virtual patient outcomes — enrollment, dropout, endpoints, safety —
  drawn from *statistical models*
- A way to evaluate **operating characteristics**
  - probabilities of success, error rates, estimation quality, timelines, resource use
- A way to make assumptions **explicit, inspectable and challengeable**


## Trial simulation is *not* :no_entry:

- A power calculation *(power is one operating characteristic among many)*
- A "virtual patient" biological or QSP simulation
- Enrollment forecasting *(though enrollment can be part of it)*
- A substitute for judgment — it prices your assumptions, it does not choose them

## Why patient-level simulation? :microscope: {.small}

::: {.fragment .keybox}

Simulation lets you inspect **individual patient-level replicates**, not just averaged operating characteristics.

:::


::: {.fragment}

An 80% probability of success tells you the design usually works. It says little about:

:::


- what the **20%** failures have in common
- which patients or analyses drive a borderline decision
- what one plausible realized trial looks like

::: {.fragment}

Operating characteristics are the *answer*. Individual replicates are the *explanation* — and that is what earns buy-in.

:::


# Regulatory guidance :scroll: {background-color="#1d5c4a"}

## Simulation as expected practice :ledger: {.small}

- **FDA** design simulations thoroughly described in *Adaptive Design Guidance*
  [@fda2019adaptive] and additional guidance in *Complex and Innovative Trial Designs
  Guidance* [@fda2020cid] and *Bayesian Methodology Guidance* [@fda2026bayesian]
- **EMA** design simulations mentioned in *Complex Clinical Trials Q&A*
  [@ec2022complex] and *Extrapolation for Pediatrics Reflection* [@ema2018extrapolation]
- **ICH** design simulation thoroughly described in *E20: Adaptive Designs For Clinical
  Trials* [@ich2025e20], including a dedicated section on conducting and reporting
  simulation studies


## What regulators are asking for :mag:

- **Pre-specification** — scenarios, decision rules and analysis models fixed before
  the data exist
- **Justification of the scenario set** — why *these* truths, and are unfavourable
  ones represented?
- **Reproducibility** — seeds, versions, code that someone else can run
- **Transparency about approximation** — Monte Carlo error, number of replicates, and
  what was *not* simulated


::: {.fragment .keybox}

Every one of these is a **software engineering** property as much as a statistical one.
That is the argument for a framework rather than a one-off script.

:::


# A running example :pill: {background-color="#1d5c4a"}

## Our design problem :chart_with_downwards_trend: [Example]{.example-tag} {.small}

::: {.fragment}

**Planning a fixed Phase 2 randomized dose-finding study in weight loss.**

:::


::: {.fragment}

| | |
|---|---|
| **Population** | Adults with obesity |
| **Endpoint** | Percent change from baseline in body weight at Week 24 |
| **Candidate doses** | 0 (placebo), 5, 10, 25, 50, 100 mg QW; MTD = 100 mg |
| **Max sample size** | <300  |
| **Allocation** | Uncertain; team's instinct is 1:1 across doses, but placebo and/or top dose could be weighted higher |
| **Dose response** | Shape uncertain; all preclinical and PK/PD data support monotonicity |
| **Analysis** | MCP-Mod or Dunnett's Test |
| **Clinically viable target** | 15% placebo-adjusted weight loss (percent change from baseline) |

:::



::: {.fragment .keybox}

Every uncertainty above is a **choice we have to make anyway** — simulation just makes
the cost of each choice visible.




From here, slides marked [Example]{.example-tag} build up the specification of this
study, component by component. By the end you will have a complete simulation plan —
without having written the simulation.

:::


# Five essential components :jigsaw: {background-color="#1d5c4a"}

## Anatomy of a trial simulation :anatomical_heart: {.small}

::: {.fragment}

```{mermaid}
%%| fig-width: 12
%%{init: {"themeVariables": {"fontSize": "18px", "edgeLabelBackground": "#f6faf8"}, "flowchart": {"htmlLabels": true, "useMaxWidth": false}}}%%
flowchart LR
  subgraph L[" "]
    direction TB
    O(<b>1. Objectives</b><br/>quantifiable targets) --> S(<b>2. Scenarios</b><br/>design x truth grid)
  end
  subgraph M[" "]
    direction LR
    D(<b>3. Data generation</b><br/>patient-level models) --> T(<b>4. Trial design</b><br/>milestones, analyses, adaptations)
  end
  E(<b>5. Evaluation criteria</b><br/>operating characteristics)
  S --> D
  T --> E
  E -.->|refine scenarios| S
  E -.->|reframe objectives| O
  style L fill:none,stroke:none
  style M fill:none,stroke:none
```

:::


::: {.fragment}

Every simulation has all five. Most disagreements are really about **one** of them — naming it keeps the discussion tractable.

Let's go through them.

:::


## 1. Objectives :goal_net: {.small}

::: {.fragment}

Turning a clinical question into something a computer can count.

:::


- Start from the **decision** the trial has to support, not from the analysis
- State the target as a **probability of an event you can observe in a replicate**
- Attach a **threshold** — a number the design has to clear to be acceptable
- Objectives under *favourable* truths and objectives under *unfavourable* truths are
  often necessary; a power target without a false-positive budget is only half an
  objective


::: {.fragment .keybox}

If you cannot write the objective as `mean(indicator_over_replicates) >= threshold`,
it is likely not yet an objective.

:::


## 1. Objectives [Example]{.example-tag} :dart: {.small}

::: {.fragment}

For our weight-loss dose-finding study:

:::


- **≥ 70%** probability of meeting the success criterion under **each positive** truth scenario
- **≤ 15%** probability of meeting the success criterion under the **negative** truth scenario


::: {.fragment .keybox}

**Success** = a **significant** dose-response signal (α = 0.10, one-sided) **and** a
selected dose whose **true** effect is at least a clinically viable effect of 15%.

:::


::: {.fragment}

Under the negative truth *no* dose reaches 15%, so the second line is a
**false-positive decision** — the probability of claiming a viable dose exists when none
does. 

:::

## 2. Scenarios: the world vs. our choices :earth_africa: {.small}

::: {.fragment}

**Scenario** = an input parameter combination we iterate across. Two types:

:::

::::: {.columns}
:::: {.column width="48%"}

::: {.fragment .keybox}
**Design scenarios** — *our choices*

Sample size, number of arms, allocation ratio, populations, decision rules, analysis
model, timing of analyses, adaptive rules.

*We control these. The output is for comparison*
:::

::::
:::: {.column width="48%"}

::: {.fragment .keybox}
**Truth scenarios** — *the world*

Assumed or plausible realities governing treatment effect and outcomes: effect sizes,
dose-response shape, variability, dropout, event rates.

*We do not control these. The output is for robustness.*
:::

::::
:::::

::: {.fragment .zoom}

The **cross-product** of design × truth scenarios defines the simulation grid.

:::


## 2. Scenarios: why the distinction matters ⚖️

- The two types answer **different questions**, and conflating them produces the classic
  mistake of optimizing a design against a single assumed truth
- **Marginalize over truths** → how robust is this design?
- **Compare across designs** → which design should we run?


::: {.fragment}

**Simulation asks:** *across the truths we consider plausible, which of our choices
performs best on the metrics we care about?*

:::


## 2. Design scenarios [Example]{.example-tag} :triangular_ruler: {.small}

::: {.fragment}

| Design parameter | Values explored |
|---|---|
| **Number of active arms** | 3 arms (25, 50, 100 mg) · 4 arms (10, 25, 50, 100 mg) · 5 arms (5, 10, 25, 50, 100 mg) |
| **Total sample size** | Fixed at 240 to start — deliberately simplifying |
| **Allocation ratio** | Equal (1:…:1) · Weighted placebo and top dose (2:1:…:1:2) |
| **Analysis model** | MCP-Mod **and** Dunnett — both computed on every replicate, so they compare methods rather than multiply the grid |

:::


::: {.fragment}

3 arm counts × 2 allocations = **6 design scenarios** — each one a trial we
could actually run.

:::


::: {.fragment}

Note that sample size is *fixed* here only for teaching. In practice it is often the first
thing that would be varied

:::


## 2. Truth scenarios [Example]{.example-tag} :crystal_ball: {.small}

::: {.fragment}

Assume placebo mean response = 0 and dose response follows a (sigmoid) Emax model:

:::


::: {.fragment}

$$f(d) = E_{\max}\,\frac{d^{\,h}}{ED_{50}^{\,h} + d^{\,h}}$$

:::


::: {.fragment}

| Scenario | $ED_{50}$ | $h$ | Shape | $E_{\max}$ *(derived)* | $f(100)$ |
|---|---|---|---|---|---|
| **Negative** | 7.5 mg | 1 | hyperbolic | 13.98 | 13% — *never clinically viable* |
| **Emax1** | 4 mg | 1 | hyperbolic | 18.72 | 18% — *true MED 25 mg* |
| **Emax2** | 20 mg | 3 | sigmoid | 18.14 | 18% — *true MED 50 mg* |
| **Emax3** | 50 mg | 3 | sigmoid | 20.25 | 18% — *true MED 100 mg* |

:::


::: {.fragment .keybox}

$E_{\max}$ is the **asymptote**, not the effect at a studied dose. If we fixed it at 18%,
every curve would land *below* 18% at 100 mg by a different amount. So we instead fix
$f(100)=18\%$ and back-solve $E_{\max} = \Delta\left[1 + (ED_{50}/d^*)^{h}\right]$.
Now the positive truths differ only in **shape**, not in what the top dose is worth.

:::


## 2. Truth scenarios, visualized [Example]{.example-tag} :chart_with_upwards_trend:

::: {.fragment}

```{r}
#| label: truth-scenarios-plot
#| fig-width: 11
#| fig-height: 4.8

candidate_doses <- c(0, 5, 10, 25, 50, 100)
dose_grid       <- seq(0, 100, length.out = 401)
target_effect   <- 15    # clinically viable placebo-adjusted effect
top_dose        <- 100   # highest studied dose (MTD)
top_effect      <- 18    # positive truths: effect delivered AT the top dose
neg_effect      <- 13    # negative truth: effect delivered AT the top dose

# Emax is the model's asymptote, not the effect at any dose we study. Back-solve it
# so that f(top_dose) == the intended top-dose effect exactly:
#   f(d) = Emax * d^h / (ED50^h + d^h)  =>  Emax = effect * (1 + (ED50 / d)^h)
emax_for <- function(effect, ed50, h, d = top_dose) effect * (1 + (ed50 / d)^h)

truth_scenarios <- tibble::tribble(
  ~scenario,  ~kind,      ~ed50, ~h,
  "Negative", "Negative",   7.5,  1,
  "Emax1",    "Positive",     4,  1,
  "Emax2",    "Positive",    20,  3,
  "Emax3",    "Positive",    50,  3
) |>
  mutate(eMax = emax_for(if_else(kind == "Positive", top_effect, neg_effect),
                         ed50, h))

curves <- truth_scenarios |>
  rowwise() |>
  reframe(scenario = scenario, kind = kind, dose = dose_grid,
          mean = sigEmax(dose_grid, e0 = 0, eMax = eMax, ed50 = ed50, h = h))

marks <- truth_scenarios |>
  rowwise() |>
  reframe(scenario = scenario, kind = kind, dose = candidate_doses,
          mean = sigEmax(candidate_doses, e0 = 0, eMax = eMax, ed50 = ed50, h = h))

pal <- c(Negative = "#c0392b", Emax1 = "#8ec4e0", Emax2 = "#2c6e9e",
         Emax3 = "#1b7837")

target_label <- tibble::tibble(kind = "Negative", dose = 0, mean = target_effect + 1.4,
                               scenario = "Negative",
                               lab = "Clinically viable effect = 15%")

ggplot(curves, aes(dose, mean, colour = scenario, linetype = kind)) +
  geom_hline(yintercept = target_effect, linetype = "dotted", linewidth = 0.8,
             colour = "grey25") +
  geom_line(linewidth = 1.2) +
  geom_point(data = marks, size = 2.2, show.legend = FALSE) +
  geom_text(data = target_label, aes(label = lab), hjust = 0, size = 4.2,
            colour = "grey25", show.legend = FALSE) +
  scale_colour_manual(values = pal, name = NULL,
                      breaks = c("Negative", "Emax1", "Emax2", "Emax3")) +
  scale_linetype_manual(values = c(Negative = "dashed", Positive = "solid"),
                        guide = "none") +
  scale_x_continuous(breaks = candidate_doses) +
  scale_y_continuous(limits = c(0, 21), breaks = seq(0, 20, 5)) +
  labs(x = "Dose (mg QW)",
       y = "Placebo-adjusted\n% weight loss at Week 24") +
  facet_wrap(~ kind, ncol = 2)
```

:::


## 2. The Scenario "grid" [Example]{.example-tag} 🔢

::: {.fragment}

| | |
|---|---|
| Design scenarios | 3 arm counts × 2 allocation schemes = **6** |
| Analysis models | MCP-Mod + Dunnett — run on **every** replicate, not extra grid rows |
| Truth scenarios | 1 negative + 3 positive = **4** |
| **Simulation grid** | 6 × 4 = **24 cells** |
| Replicates per cell | 1,000 |
| **Total virtual trials** | **24,000** |

:::


::: {.fragment .keybox}

This is a *deliberately simplified* example, and it is already 24 cells. Adding
four sample sizes makes it 96 — this blows up quickly. In code, represent it as a
dataframe with one row per scenario and use tools like `tidyr::expand_grid()` to
add and filter scenarios easily.

:::


## 3. Data generation models :game_die:

::: {.fragment}

Statistical models are then used to simulate patient data, including:

:::


- **Single / multiple endpoints** of varying type: continuous, binary, count, time-to-event
- **Longitudinal outcomes** — correlated errors or random effects
- **Trial mechanics** — stochastic enrollment, dropout, missing visits, administrative
  censoring


::: {.aside}

A simulation study is a modeling exercise in disguise

:::


## 3. Data generation model [Example]{.example-tag} :game_die:

::: {.fragment}

For our study, deliberately the simplest thing that could work:

:::


::: {.fragment}

$$Y_{ij} \sim \mathrm{N}\left(f(d_j),\ \sigma^2\right), \qquad \sigma = 10$$

:::


- Percent change from baseline in weight is generated **directly at the patient level**
- $f(d_j)$ is the sigmoid EMAX mean from the active truth scenario
- Fixed variance across arms; no dropout, no covariates

::: {.fragment .keybox}

Note how little this model contains — and note that it is still a *complete, runnable
specification*. Every elaboration you would want (dropout, a Week 12 interim readout, a
correlated safety endpoint) is an addition to the code - not switching tools/modules

:::


## 4. Trial design :building_construction: {.small}

::: {.fragment}

Everything that happens between "patients exist" and "we have a decision."

:::


::: {.fragment}

| | |
|---|---|
| **Milestones / analyses** | When do we look, and what data are available at each look? |
| **Analysis models** | What we actually fit — need not (and deliberately should not, when testing robustness) match the data generation model |
| **Adaptive elements** | Interim analyses, futility and efficacy stopping, arm dropping or addition, response-adaptive randomization, sample size re-estimation |
| **Decision rules** | The deterministic function that turns an analysis into a Go, a No-Go, or a selected dose |

:::


::: {.fragment .keybox}

The decision rule is the part teams most often leave vague, and the part simulation is
least able to guess.

:::


## 4. Trial design [Example]{.example-tag}: significance testing :test_tube: {.small}

::: {.fragment}

Single final analysis at Week 24. Two analysis models, both at **α = 0.10** (one-sided):

:::


::: {.fragment}

**MCP-Mod** — candidate model set:

:::


::: {.fragment}

| Candidate | Form | Parameters |
|---|---|---|
| C1 | hyperbolic Emax | $ED_{50}$ = 5 |
| C2 | hyperbolic Emax | $ED_{50}$ = 25 |
| C3 | sigmoid Emax | $ED_{50}$ = 5, $h$ = 2 |
| C4 | sigmoid Emax | $ED_{50}$ = 25, $h$ = 2 |
| Linear | linear | — |
| Exponential | exponential | $\delta$ = 25 |

:::


::: {.fragment}

*Candidate labels (C1–C4) are deliberately distinct from the truth-scenario names
(Emax1–Emax3): these are the shapes we **fit**, not the shapes the world takes.*

:::


::: {.fragment}

**Dunnett's Test** — all active doses vs. placebo.

:::


## 4. Trial design [Example]{.example-tag}: candidate models :ocean: {.small}

::: {.fragment}

```{r}
#| label: mcpmod-candidates
#| fig-width: 11
#| fig-height: 4.2

candidate_models <- Mods(
  emax        = c(5, 25),                    # C1, C2  (hyperbolic)
  sigEmax     = rbind(c(5, 2), c(25, 2)),    # C3, C4  (sigmoid, h = 2)
  linear      = NULL,
  exponential = 25,
  doses       = candidate_doses,
  placEff     = 0,
  maxEff      = top_effect
)

cand_labels <- c(emax1 = "C1  Emax (ED50 = 5)",  emax2 = "C2  Emax (ED50 = 25)",
                 sigEmax1 = "C3  sigEmax (ED50 = 5, h = 2)",
                 sigEmax2 = "C4  sigEmax (ED50 = 25, h = 2)",
                 linear = "Linear", exponential = "Exponential (delta = 25)")

cand_curves <- getResp(candidate_models, doses = dose_grid) |>
  as.data.frame() |>
  mutate(dose = dose_grid) |>
  pivot_longer(-dose, names_to = "model", values_to = "mean") |>
  mutate(model = factor(cand_labels[model], levels = unname(cand_labels)))

ggplot(cand_curves, aes(dose, mean)) +
  geom_hline(yintercept = target_effect, linetype = "dotted", colour = "grey40") +
  geom_line(linewidth = 1.1, colour = "#2c6e9e") +
  geom_vline(xintercept = candidate_doses, colour = "grey85", linewidth = 0.3) +
  scale_x_continuous(breaks = c(0, 25, 50, 100)) +
  scale_y_continuous(breaks = seq(0, 20, 10)) +
  labs(x = "Dose (mg QW)", y = "Standardized\nmean response") +
  facet_wrap(~ model, ncol = 3)
```

:::


::: {.fragment}

Candidates are **shapes**, not truths. They span the plausible dose-response space so
some candidate is close to whatever the world does. The dotted line is the 15% target;
the grey verticals are the candidate doses.

:::


## MCP-Mod vs. Dunnett's test :boxing_glove: {.small}

::: {.fragment}

Two potential answers to "is there a dose-response signal?"

:::


::: {.fragment}

| | **Dunnett** [@dunnett1955] | **MCP-Mod** [@bretz2005; @pinheiro2014] |
|---|---|---|
| Question asked | Is *any* dose better than placebo? | Is there a dose-response *signal*, and what *shape*? |
| Uses dose ordering | No — arms are exchangeable labels | Yes — borrows strength across the dose range |
| Multiplicity over | Treatment arms | Candidate models |
| Estimation | Based on observed data | Based on fitted dose-response model |
| Cost of misspecification | None — model-free | Power loss if no candidate is close |

:::


::: {.fragment .keybox}

MCP-Mod is well established in this setting - dose-response modelling improves estimation of dose-response
curves and can be advantageous for decision-making with more doses.

:::


## 4. Trial design [Example]{.example-tag}: dose selection :round_pushpin: {.small}

::: {.fragment .keybox}

**MED** (minimally efficacious dose) = the lowest dose estimated to have a
placebo-adjusted effect **> 15%**.




*MCP-Mod:* estimate MED from the fitted curve. *Dunnett:* estimate it from observed arm means.

:::


::: {.fragment}

**"Acceptable dose selection"** = selecting an MED that is either

:::


1. **correct** for the scenario in question — the ideal — **or**
2. **incorrect, but still has at least a 15% treatment effect in truth**

::: {.fragment}

Why the second clause? Phase 2 should not be scored as a failure just because it misses the
*exact* MED. If it lands on a different dose that still delivers clinically meaningful efficacy,
that is still a **good outcome**.

:::


::: {.fragment .success-box}

**Success** = statistical significance **and** acceptable dose selection.

:::


## 5. Evaluation criteria :bar_chart:

::: {.fragment}

Turning replicates into numbers a team can act on.

:::


- Metrics come **directly from the objectives** — if a metric does not map to an
  objective, ask whether it needs to be computed at all
- Add **diagnostic** metrics beyond the decision metrics — they explain *why* a design
  performs the way it does
- Keep replicate-level output (or a chosen subset) — you will want it the first
  time an aggregate surprises you


## 5. Evaluation criteria [Example]{.example-tag} :bar_chart:

::: {.fragment}

**From the objectives:**

:::


- **Success rate** under each positive truth scenario — target ≥ 70%
- **False-positive rate** under the negative truth scenario — target ≤ 15%

::: {.fragment}

**Metrics we could add, once the objectives are met:**

:::


- **MSE or bias of the estimated dose-response** across the candidate dose range

::: {.fragment}

Neither is in the objectives, but they explain *why* two designs differ — and
dose-response estimation matters beyond the go/no-go decision, since real dose
selection weighs factors other than efficacy.

:::


## The components [Example]{.example-tag} :clipboard: {.small}

::: {.fragment}

| Component | Our study |
|---|---|
| **Objectives** | ≥ 70% success under positive truths; ≤ 15% false positive under the negative truth |
| **Scenarios** | 6 design × 4 truth = 24-cell grid; both analysis models on every replicate |
| **Data generation** | Patient-level data generation model using sigmoid EMAX model |
| **Trial design** | Single final analysis; MCP-Mod or Dunnett at α = 0.10; MED rule; success = significance + acceptable dose selection |
| **Evaluation** | Success rate, false-positive rate, replicate-level inspection |

:::


::: {.fragment}

That is a mostly complete simulation plan for an initial simulation run. We have not written the simulation — but there is
now nothing left to argue about except the code.

:::


# The R-based computational framework :computer: {background-color="#1d5c4a"}

## Why R and why we are being opinionated :rocket: {.small}

::: {.incremental}
- **The statistical models already live there.** `DoseFinding`, `rpact`, `brms`,
  `mmrm`, survival tooling, copula tooling — the analysis models you want to simulate
  are R packages, and calling them beats reimplementing them.
- **Statisticians are already using R.** The people you are consulting with are using
  and likely can provide R code to get started with.
- **R is very flexible through functional programming.** It is easy to update matching
  the quick iterative process of designing a trial.  If intentionally developed, it is
  simple to adapt.
- **Adding and subtracting scenarios is easy.** A simulation grid is a data frame;
  a simulation is a function mapped over its rows. That one idea does most of the work.
- **R is becoming more standard in pharma** 

:::


::: {.fragment}

The moment the team asks something reasonable — *"what about an extra endpoint at a
Week 12 interim?"* — a packaged tool either accommodates it or it does not. R always
does.

:::


::: {.fragment .keybox}

This course teaches a specific R-based approach. It is not a survey of options.

:::


# Over to Part 2 :handshake: {background-color="#1d5c4a"}

## Where Saumil picks up :arrow_right:

::: {.fragment}

Part 1 gave you the five components and a fully specified study. Part 2 writes them.

:::
