```{r}
#| label: setup-p3
#| include: false

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", dev = "svglite", dev.args = list(bg = "transparent"))

library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(
  theme_minimal(base_size = 15) +
    theme(panel.grid.minor = element_blank(),
          legend.position = "right",
          plot.background = element_rect(fill = "transparent", colour = NA),
          panel.background = element_rect(fill = "transparent", colour = NA))
)

teal <- "#1d5c4a"; teal_mid <- "#2a7a63"; teal_lt <- "#8fc4b3"; warn <- "#c0392b"
ink  <- "#1e2a27"; ink_mu <- "#6b7770"

# Truth colours are inherited from part 1 (same truth, same colour all course).
# Hue family encodes ED50, lightness encodes h -- composite encoding, so the
# series are not distinguished by hue alone.
pal_truth  <- c(Negative = "#c0392b", Emax1 = "#8ec4e0", Emax2 = "#2c6e9e",
                Emax3 = "#1b7837")
# Two methods: one hue, separated by a large lightness step.
pal_method <- c(Dunnett = teal_lt, `MCP-Mod` = teal)

# Scenario definitions are sourced from the same file the simulation uses, so
# every number quoted in this deck is derived, not transcribed -- it cannot
# drift when the scenarios are re-tuned. (This only builds data frames; it runs
# no simulation.)
source("R/part-3/1-scenarios.R")

N_STUDY <- unique(design_scenarios$sample_size)
TOP_EFF <- top_effect
NEG_EFF <- neg_effect
FULL_DOSES <- arm_sets$doses[[which.max(arm_sets$n_active)]]

# True effect at every dose, and the true MED, for each truth scenario.
truth_meds <- sapply(seq_len(nrow(truth_scenarios)), function(i) {
  t <- truth_scenarios[i, ]
  true_med_for(FULL_DOSES, t$eMax, t$ed50, t$h)
})

# The analysis on these slides IS the exercise solution -- we source it, so what
# is on screen and what an attendee can open are guaranteed to be the same
# thing. It reads the pre-run results file and simulates nothing.
have_results <- file.exists("data/p3-results.rds")
if (have_results) {
  # Guard against a results file left over from an earlier scenario set: if the
  # truths it was run under no longer match the current definitions, treat it as
  # absent rather than silently plotting numbers from a study we no longer run.
  .stored <- readRDS("data/p3-results.rds")
  have_results <- setequal(unique(.stored$replicates$truth), truth_scenarios$truth)
  rm(.stored)
}
if (have_results) {
  invisible(capture.output(suppressMessages({
    source("R/part-3/7-exercise.R")
    source("R/part-3/8-exercise-solution.R")
  })))
}
```

# What we are doing :dart: {background-color="#1d5c4a"}

## Where we are :round_pushpin: {.smaller}

| | Time | Session | | |
|:--:|---|---|---|---|
| **1** | 1:30 – 2:30 PM | **Overview of Trial Simulations**<br/>*Mitch · 60 min* | Motivation · modern capabilities · regulatory guidance<br/>The five essential components, on one running example | done |
| | 2:30 – 2:40 PM | 10 minute break | | |
| **2** | 2:40 – 3:40 PM | **Programming Key Elements in R**<br/>*Saumil · 60 min* | Objectives · scenarios · data generation<br/>Complex designs · evaluation · parallelization | done |
| | 3:40 – 3:50 PM | 10 minute break | | |
| **3** | 3:50 – 5:00 PM | **Hands-on Exercise**<br/>*Mitch · flexible* | Build one design end to end<br/>Compare designs across the full study | **← you are here** |


## The study, already run [Example]{.example-tag} :checkered_flag: {.smaller}

::: {.fragment}

| | |
|---|---|
| **Endpoint** | % change from baseline in body weight at Week 24, $\sigma = 10$ |
| **Doses** | 0 (placebo), 5, 10, 25, 50, 100 mg QW |
| **Sample size** | `r if (exists("N_STUDY")) N_STUDY else 240` total, the same for every design |
| **Designs** | 3 / 4 / 5 active arms × equal or weighted allocation = **6** |
| **Truths** | 1 negative + 3 positive Emax curves = **4**, with true MEDs of 25 / 50 / 100 mg |
| **Analyses** | MCP-Mod **and** Dunnett, both on every replicate, $\alpha = 0.10$ |
| **Success** | statistical significance **and** acceptable dose selection |

:::


::: {.fragment .zoom}

6 × 4 = **24 cells** × 1,000 replicates = **24,000 virtual trials** — already run.

:::


## The code is yours — just not today :gift: {.smaller}

::: {.fragment}

```{mermaid}
%%| fig-width: 12
%%{init: {"themeVariables": {"fontSize": "18px"}, "flowchart": {"htmlLabels": true, "useMaxWidth": false}}}%%
flowchart LR
  S["<b>1-scenarios.R</b><br/>design x truth grid"] --> D["<b>2-datagen.R</b><br/>population generators"]
  D --> T["<b>3-design.R</b><br/>triggers + analyses + MED"]
  T --> E["<b>4-evaluate.R</b><br/>objectives + OCs"]
  E --> R["<b>6-runsim.R</b><br/>runs the grid via run_one()"]
```

:::


::: {.columns}
::: {.column width="55%"}

**Four numbered components**, then the driver that runs them — the same five
ideas from Part 1, in the same rxsim vocabulary Saumil used in Part 2.
`run_one()` is the helper in `R/part-3/5-run-one.R`, exactly the shape of Part 2's
`simulate_scenario()`.

:::
::: {.column width="45%"}

::: {.fragment .keybox}
**We are not walking through this today.** It is in the repo, it runs, and it
is yours to pull apart afterwards.

Today we work with what it **produced**.
:::

:::
:::


## The four truth curves :chart_with_upwards_trend: {.smaller}

::: {.fragment}

```{r}
#| label: truth-curves
#| echo: false
#| fig-width: 11
#| fig-height: 4.2

dose_grid <- seq(0, 100, length.out = 401)

curve_at <- function(d) {
  do.call(rbind, lapply(seq_len(nrow(truth_scenarios)), function(i) {
    t <- truth_scenarios[i, ]
    data.frame(truth = t$truth, kind = t$kind, dose = d,
               mean = DoseFinding::sigEmax(d, e0 = 0, eMax = t$eMax,
                                           ed50 = t$ed50, h = t$h))
  }))
}
curves <- curve_at(dose_grid)
marks  <- curve_at(FULL_DOSES)

target_label <- data.frame(kind = "Negative", dose = 0, mean = Delta + 1.3,
                           truth = "Negative",
                           lab = sprintf("Clinically viable effect = %d%%", Delta))

ggplot(curves, aes(dose, mean, colour = truth, linetype = kind)) +
  geom_hline(yintercept = Delta, linetype = "dotted", linewidth = 0.8,
             colour = "grey25") +
  geom_line(linewidth = 1.2) +
  geom_point(data = marks, size = 2.2, show.legend = FALSE) +
  geom_text(data = target_label, aes(label = lab), hjust = 0, size = 4.2,
            colour = "grey25", show.legend = FALSE) +
  scale_colour_manual(values = pal_truth, name = NULL,
                      breaks = truth_scenarios$truth) +
  scale_linetype_manual(values = c(Negative = "dashed", Positive = "solid"),
                        guide = "none") +
  scale_x_continuous(breaks = FULL_DOSES) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  labs(x = "Dose (mg QW)", y = "Placebo-adjusted\n% weight loss at Week 24") +
  facet_wrap(~ kind, ncol = 2)
```

:::


::: {.fragment}

The **true MED** — the lowest studied dose truly worth ≥ 15% — is
`r paste(sort(truth_meds), collapse = " / ")` mg for Emax1 / Emax2 / Emax3, and
none for the negative truth. That is the target each trial is trying to recover.

:::


::: {.fragment}

Every positive truth reaches **`r if (exists("TOP_EFF")) TOP_EFF else 18`%** at
100 mg. They differ only in **where they cross the dotted line** — that is the
MED problem each design has to solve.

:::


# Your turn :raising_hand: {background-color="#1d5c4a"}

## What you have: one row per trial :toolbox: {.smaller}

::: {.fragment}

Open **`R/part-3/7-exercise.R`**. It loads the completed study, plus the same
helpers used to score it.

:::


::: {.fragment}

`r if (have_results) sprintf("**%s rows x %d columns** -- %d designs x %d truths x %s replicates. One design, all four truths, first two replicates:", format(nrow(reps), big.mark = ","), ncol(reps), dplyr::n_distinct(reps$design), dplyr::n_distinct(reps$truth), format(results$n_rep, big.mark = ","))`

:::


::: {.fragment}

```{r}
#| label: reps-preview
#| echo: false
if (have_results) {
  reps |>
    dplyr::filter(design == design[1], replicate <= 2) |>
    dplyr::select(design, truth, replicate,
                  dunn_sig, dunn_med, mcp_sig, mcp_med, obs_d100) |>
    dplyr::mutate(obs_d100 = round(obs_d100, 1)) |>
    knitr::kable(align = "llrrrrrr")
} else {
  knitr::kable(data.frame(note = "Run R/part-3/6-runsim.R to populate."))
}
```

:::


::: {.fragment}

*8 of 31 columns shown — the rest are `obs_d*` / `mod_d*`, the observed and
fitted effect at each studied dose.*

:::


## What you have: columns and helpers :toolbox: {.smaller}

::: {.columns}
::: {.column width="54%"}

`reps` — **one row per simulated trial**



::: {.fragment}

| Column | |
|---|---|
| `design`, `truth` | which cell |
| `dunn_sig`, `mcp_sig` | signal detected? |
| `dunn_med`, `mcp_med` | dose selected (`NA` = none) |
| `dunn_min_p`, `mcp_min_p` | the p-values |

:::

:::
::: {.column width="46%"}

`results$oc` — the same thing pre-summarised.

::: {.fragment .keybox}
Do it from `reps` first, then use `results$oc` to check.
:::

::: {.fragment}

From `4-evaluate.R`:

```r
true_effect(truth, dose)     # what is TRUE there
acceptable(truth, med)       # was the choice good?
is_success(truth, sig, med)  # signal AND acceptable
claimed_viable(sig, med)     # claimed a viable dose?
mc_se(p, n_rep)              # is a difference real?
```

:::


::: {.fragment .keybox}

That file is the **single definition** of success — the stored results used
these same functions, so your numbers are directly comparable.

:::
:::

:::

## The question :question: {.smaller}

::: {.fragment .success-box}

Which design should we run, and why?

:::


::: {.columns}
::: {.column width="50%"}
1. Does **any** design meet both objectives?
2. Every design has a **binding truth** — which, and is it the same for all six?
3. Does studying **more doses** help? Predict before you compute.
4. **MCP-Mod vs Dunnett** — same trials, so it is a paired comparison

:::
::: {.column width="50%"}
5. Go to the **replicates**: what do the failures have in common?
6. Make **one figure** you would put in front of a design team

::: {.fragment .keybox}

Do the ones you find interesting, not all six.




Stuck? `R/part-3/8-exercise-solution.R` produces exactly what comes next.

:::
:::

:::


# What we found :mag: {background-color="#1d5c4a"}

## Does any design meet the objectives? :goal_net: {.smaller .scrollable}

::: {.keybox}

**Reminder — the two objectives from Part 1:**

**A.** `r if (have_results) sprintf("≥ %.0f%%", 100 * success_target)` success under
**each** positive truth, where success = a signal detected *and* an acceptable dose
selected.

**B.** `r if (have_results) sprintf("≤ %.0f%%", 100 * claim_target)` chance of claiming a
viable dose under the **negative** truth, where no dose is viable.

:::

::: {.fragment}

```{r}
#| label: verdict
#| echo: false
if (have_results) {
  verdict |>
    dplyr::transmute(Design = design, Method = method,
                     `Worst success` = round(min_success, 3),
                     `Binding truth` = binding_truth,
                     `Neg. claim` = round(claim_rate, 3),
                     A = ifelse(meets_A, "yes", "**no**"),
                     B = ifelse(meets_B, "yes", "**no**"),
                     Both = ifelse(meets, "**PASS**", "fail")) |>
    knitr::kable(col.names = c("Design", "Method", "Worst success",
                               "Binding truth", "Neg. claim",
                               sprintf("A (≥%.2f)", success_target),
                               sprintf("B (≤%.2f)", claim_target), "Both"))
} else {
  knitr::kable(data.frame(note = "Run R/part-3/6-runsim.R to populate."))
}
```

:::


::: {.fragment .zoom}

`r if (have_results) sprintf("**%d of %d** meet both — and the four that fail do not all fail the *same* objective.", sum(verdict$meets), nrow(verdict))`

:::


## Both objectives at once — on average 🎯 {.smaller}

::: {.fragment}

```{r}
#| label: oc-mean
#| echo: false
#| fig-height: 4.0
if (have_results) fig_oc else ggplot() + theme_void()
```

:::


::: {.fragment}

Read it like an ROC curve: false positives across, success up, **good corner =
top-left**. The shaded box is where both objectives hold.

:::


## Zoom: the designs that pass :mag_right: {.smaller}

::: {.fragment}

```{r}
#| label: oc-zoom
#| echo: false
#| fig-height: 4.0
if (have_results) fig_oc_zoom else ggplot() + theme_void()
```

:::


::: {.fragment .smaller}

```{r}
#| label: oc-zoom-note
#| echo: false
#| results: asis
if (have_results) {
  v  <- verdict |> dplyr::filter(meets)
  ms <- range(v$mean_success)
  cat(sprintf(
"**Every design that passes uses MCP-Mod** — Dunnett misses the claim-rate objective at all six. Among the survivors, adding doses lowers mean success a little (%.2f to %.2f) but all stay above %.2f, so **3, 4 and 5 arms remain viable**.",
    ms[1], ms[2], success_target))
}
```

:::


## The same designs — worst case :cloud_with_rain: {.smaller}

::: {.fragment}

```{r}
#| label: oc-min
#| echo: false
#| fig-height: 3.7
if (have_results) fig_oc_min else ggplot() + theme_void()
```

:::


::: {.fragment}

This is the stricter test: a design has to clear the bar under **every**
plausible truth, not just on average.

:::


::: {.fragment .smaller}

**This is the figure you put in front of a design team** — every design, both
objectives, judged on its worst truth. (That was exercise question 6.)

:::


## Zoom: judged on the worst truth :mag_right: {.smaller}

::: {.fragment}

```{r}
#| label: oc-zoom-min
#| echo: false
#| fig-height: 4.0
if (have_results) fig_oc_min_zoom else ggplot() + theme_void()
```

:::


::: {.fragment .smaller}

```{r}
#| label: oc-zoom-min-note
#| echo: false
#| results: asis
if (have_results) {
  v  <- verdict |> dplyr::filter(meets)
  ms <- range(v$min_success)
  cat(sprintf(
"**The same MCP-Mod designs still pass** under their worst truth: %.2f to %.2f against %.2f, so **3, 4 and 5 arms remain viable**.",
    ms[1], ms[2], success_target))
}
```

:::


## Success under every positive truth :hotsprings: {.smaller}

::: {.fragment}

```{r}
#| label: heatmap
#| echo: false
#| fig-height: 4.3
if (have_results) fig_heatmap else ggplot() + theme_void()
```

:::


::: {.fragment}

Grey is exactly the `r if (have_results) sprintf("%.0f%%", 100*success_target)` objective:
**red misses it, teal clears it**. Emax1 and Emax2 are comfortable almost
everywhere — **Emax3 is where every design is weakest**.

:::


## Estimated effect at the true MED :violin: {.smaller}

::: {.fragment}

```{r}
#| label: violin
#| echo: false
#| fig-height: 4.2
if (have_results) fig_violin else ggplot() + theme_void()
```

:::


::: {.fragment}

```{r}
#| label: violin-note
#| echo: false
#| results: asis
if (have_results) {
  sds <- est |>
    dplyr::group_by(panel, n_active) |>
    dplyr::summarise(sd = sd(estimate), .groups = "drop") |>
    tidyr::pivot_wider(names_from = n_active, values_from = sd,
                       names_prefix = "a")
  cat(sprintf(
"Each violin is the MCP-Mod estimate at that truth's true MED, across trials that succeeded. **Adding doses does not cost you precision here** — going from 3 to 5 active arms means fewer patients per dose, yet the spread is flat or *tighter* (SD %s → %s under %s, %s → %s under %s). MCP-Mod fits a curve across all doses, so the extra dose-response information offsets the smaller per-arm sample.",
    sprintf("%.2f", sds$a3[1]), sprintf("%.2f", sds$a5[1]),
    sub("  —.*", "", sds$panel[1]),
    sprintf("%.2f", sds$a3[2]), sprintf("%.2f", sds$a5[2]),
    sub("  —.*", "", sds$panel[2])))
}
```

:::


## Summary — my thoughts :thought_balloon: {.smaller}

1. **There is no single right design here.** Choosing one needs context this
   simulation does not have — cost, timelines, what the team can run. In
   practice this is an iteration, not a verdict.
2. **What the numbers did say.** More doses with fewer patients each did
   **not** meaningfully hurt the decision — the first objection to a wider
   design — and **MCP-Mod beat Dunnett on both objectives at once**.
3. **Two operating characteristics down, many to go.** Bias and MSE of the
   fitted curve, the distribution of selected doses, duration, stricter rules —
   each is another column in the same table, not a new tool.


::: {.fragment .zoom}

The iteration *is* the evidence that the design was chosen, not just arrived at.

:::


# Taking this further :rocket: {background-color="#1d5c4a"}

## What we didn't touch today :telescope: {.smaller}

::: {.fragment}

Three doors we walked past — all openable with what you already have:

:::


::: {.incremental}
- **Random truth scenarios**
  - draw truths from a distribution — say, a posterior — instead of fixing four points
  - the grid just gets one draw per replicate
- **Adaptivity**
  - Part 2's interim was informational
  - the same trigger + decision-rule machinery can stop early or drop an arm
- **Richer data generation**
  - multiple endpoints, copulas, and longitudinal models plug into the same generator
  - rxsim already works with longitudinal generators

:::


## Scaling: the grid is a data frame :stopwatch: {.smaller}

::: {.fragment}

rxsim runs one trial per core, so you parallelize **across scenario rows** —
the grid is already the unit of work.

:::


- **This study**: `r if (have_results) nrow(scenarios) else 24` cells ×
  `r if (have_results) format(results$n_rep, big.mark = ",") else "1,000"`
  replicates ≈ **`r if (have_results) format(nrow(reps), big.mark = ",") else "24,000"` virtual trials** —
  100 minutes across 6 `furrr` workers on a laptop
- **Production**: [`targets`](https://books.ropensci.org/targets/) makes each
  cell a target — it caches results and re-runs **only what changed**, which
  matters when the tenth question is "what if we add a dose?"
- `targets` + `crew` extends the same pipeline onto an HPC without rewriting
  the simulation — **extremely well equipped for exactly this kind of
  computation**. Worth watching:
  [R in Pharma talk on `targets` + `crew`](https://www.youtube.com/watch?v=cyF2dzloVLo&list=PLMtxz1fUYA5C0csy8-wTfFmRYQEB30vr3)


::: {.fragment .keybox}

The reason this scales is that nothing above is simulation-specific. A grid, a
function over its rows, and a cache.

:::


## How many simulation replicates? :100: {.smaller}

::: {.fragment}

Monte Carlo error is a design choice, not a detail. Near our
`r if (have_results) sprintf("%.2f", success_target) else "0.70"` objective:

:::


::: {.fragment}

```{r}
#| label: mc-table
#| echo: false
if (have_results) {
  n <- c(1000, 10000, 100000)
  data.frame(
    `Replicates per cell` = ifelse(n == results$n_rep,
                                   paste0(format(n, big.mark = ","), "  (today)"),
                                   format(n, big.mark = ",")),
    `Monte Carlo SE` = sprintf("±%.4f", mc_se(success_target, n)),
    check.names = FALSE) |>
    knitr::kable(align = "lr")
}
```

:::


::: {.fragment .keybox}

**How I choose it in practice**



- **~1,000** while *exploring* designs — enough to see the shape of the answer
- **≥ 10,000** once you are comparing a short list, and for the final design
- **100,000** for a single design is cheap, and worth it when the number is
  going into a submission

:::


::: {.fragment}

Or pick it from the **decision**: how big a difference between two designs
would actually change what you run? Resolve that, and stop.

:::


## Validation: simulation code is software :white_check_mark: {.smaller}

::: {.fragment .smaller}

Each component has a testable contract; they are easier to test than the whole
simulation at once.

:::

::: {.smaller}
- **Data generation**
  - large-n check that each dose hits its intended mean and SD
- **The MED rule**
  - hand-built effect vector with a known answer
  - `NA` when nothing clears Δ
- **The scenario grid**
  - size and labels are what you think they are *(6 × 4 = 24)*
:::

::: {.fragment .smaller}

`testthat` on these components, run automatically, is the difference between a
simulation you believe and one you merely hope about.

:::

::: {.fragment .keybox .smaller}

**Double programming** belongs here too — on the **custom rules**, not the
packages. The MED rule and success criterion carry the decision, and an error
there is silent.

:::


## Using AI to write simulation code :robot: {.smaller}

::: {.fragment .keybox}

A personal view — opinion, not guidance. Simply what has worked for me.

:::


::: {.fragment}

It is a **loop**, not a prompt:

:::


::: {.incremental}
1. **Ground it in code you already trust** — point it at the packages and
   simulations you already use, never a blank page
2. **Write detailed instructions** — conventions, terminology, what "good" looks
   like here; most bad output traces back to a vague ask
3. **Review everything, and be opinionated** — the first draft is a starting
   point; impose your own naming, structure, functions
4. **Feed the reviewed code back** — today's reviewed output becomes tomorrow's
   reference, and the conventions compound
:::


::: {.fragment .success-box}

Everything you have seen today was written this way

:::


# Wrap-up :tada: {background-color="#1d5c4a"}

## What we built :hammer_and_wrench: {.smaller}

- A **scenario grid** you extend by adding rows, not by editing simulation code
- A **data generation model** that is one function, replaceable without
  touching anything else
- A **trial design** where two analysis methods compete on identical trials
- A **success criterion** that encodes a clinical judgment, not just a p-value
- **Operating characteristics** that answer the question the team actually
  asked


::: {.fragment .success-box}

Every next question is a new row, not a new tool.

:::


## Today is not entirely about the package :package: {.smaller}

::: {.fragment}

Nothing today depended on a particular tool. What we are actually proposing is
a **shared way of expressing a trial**:

:::


- a scenario grid that is a **data frame**
- a simulation that is a **function mapped over its rows**
- **replicates kept**, not just the averages
- the five components **written and documented**

::: {.fragment .keybox}

Tools will change — `rxsim` may not be what you end up using. The **framework**
is what we are asking you to take home.

:::


## What we get if enough of us work this way :seedling: {.smaller}

::: {.incremental}
- **Better trials** — when one more design question costs *one more row in a data
  frame*, you ask the questions you would otherwise have talked yourself out of.
  Today's example answered thirty.
- **Designs that can actually be reviewed** — every design considered, every truth
  assumed, every objective, in code and under version control, so a reviewer can
  put a finger on the assumption they disagree with.
- **Transparency that is real** — submitting executable code, including what you
  **rejected** and why, beats describing what your code did.
:::


::: {.fragment .success-box}

The goal is not a shared package. It is a common language for design.

:::
